DATA 621 - HW1 - Regression Model

### 2020-09-10 HW1/Regression Model - Deokinanan.rmd
edited for pdf format

---
title: "DATA 621 - HW1 - Regression Model"
author: "Samantha Deokinanan"
date: "September 3, 2020"
output: html_document
---

***
#### Task
***

@@

***
#### The dataset
***

@@

```{r message=FALSE, warning=FALSE}
# The required R packages
library(tidyverse)
library(psych)
library(olsrr)
library(rcompanion)
library(mice)
library(VIM)
library(mctest)
library(caret)

# Load training data set
train_df = read.csv("C:/Users/Deokinanan/Desktop/2019 -2021 CUNY SPS/FALL 2020/DATA 621/WEEK 3/moneyball-training-data.csv")
train_df = subset(train_df, select = -c(INDEX))
```

***
#### Data Transformation, Outliers and Missing Data
***

@@

```{r}
# Quick Exploration
sapply(train_df, describe)

@@

Next, the outlier plots revealed that there are a few extreme values that can influence the analysis. Because the objective is so create a multivariate regression model, declaring an observation as an outlier based on a just one feature could lead to unrealistic inferences. There Cook's distance is use to decide if an individual entity is an extreme value or not.

Cook’s distance is a measure computed with respect to a given regression model and therefore is impacted only by the predictors included in the model. It computes the influence exerted by each data point on the response variable. Fitting the full model during this exploratory stage, the process remove the most influential points. Note from the plots that all influential observations are not necessarily outliers. Least than 3% of the data was removed.

```{r}
# Outlier Plots
model = lm(TARGET_WINS ~ ., data = train_df)
ols_plot_diagnostics(model)

@@

# New diagnostic plots
model_new = lm(TARGET_WINS ~ ., data = train_df)
ols_plot_diagnostics(model_new)
```

@@

```{r}
aggr(train_df, col = c('darkseagreen2','darkslateblue'), numbers = TRUE, sortVars = TRUE, oma = c(10,5,5,3), labels = names(train_df), cex.axis = 0.8, gap = 3, axes = TRUE, Prop = TRUE, ylab = c("Proportion of Missing Data", "Combination"))

train_df.clean = complete(mice(data = train_df, method = "pmm", maxit = 5, seed = 525, print = FALSE), 3)
sapply(train_df.clean, describe)
```

***
#### Initial Tests 
***

@@

```{r}
# Data Transformation
train_df.clean$TEAM_BATTING_1B = train_df.clean$TEAM_BATTING_H - (train_df.clean$TEAM_BATTING_2B + train_df.clean$TEAM_BATTING_3B + train_df.clean$TEAM_BATTING_HR)
train_df.clean = subset(train_df.clean, select = -c(TEAM_BATTING_H))

@@

```{r, fig.width=7, fig.height=9}
# Correlation
corr = round(cor(train_df.clean), 2)
corr = cbind(names = rownames(corr), data.frame(corr, row.names=NULL))
g.corr = gather(as.data.frame(corr), key, value, 2:length(corr), na.rm = TRUE)

ggplot(data = g.corr, aes(key, names, fill = value)) + geom_tile(color = "white") +
 scale_fill_gradient2(low = "darkseagreen2", high = "darkslateblue", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", name="Pearson\nCorrelation") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 12, hjust = 1)) +
 coord_fixed() + geom_text(aes(key, names, label = value), color = "black", size = 3) 
```

Lastly, because a robust regression process will be performed, higher order polynomials variables were introduced into the full model.

```{r}
stepdata = cbind(train_df.clean,sapply(train_df.clean[2:length(train_df.clean)], function(x) x^2))
names(stepdata) = make.unique(names(stepdata), sep = "_")
```

***
#### Regression Model #1
***

A stepwise variable selection model is conducted to determine what are the variables that can help predict the number of wins for the team. The stepwise variable selection allows variables to be added one at a time to the model, as long as the F-statistic is below the specified $\alpha$, in this case $\alpha = 0.05$. However, variables already in the model do not necessarily stay in. The steps evaluate all of the variables already included in the model and remove any variable that has an insignificant F-statistic. Only after this test ends, is the best model found, that is when none of the variables cstepan be excluded and every variable included in the model is significant.

@@

After the steps, the final model resulted below, with $adj. R^2 = 0.42$, suggesting that this model accounts for nearly 42% of the variation in the dependent variable with the independent variables, which is acceptable as a good model. Within the method of `lmStepAIC`, AIC (Akaike Information Criteria) quantifies the amount of information loss due to simplification. That is, based on the AIC, the final model outputted is the simplest model without impacting much on the performance.

```{r  message=FALSE, warning=FALSE, paged.print=FALSE}
# Set up repeated k-fold cross-validation
set.seed(525)
train.control = trainControl(method = "cv", number = 10)

@@

# Model accuracy
step.model$results

@@

# Summary of the model
summary(step.model$finalModel)
```

***
#### Prediction
***

@@

```{r message=FALSE, warning=FALSE}
# load the test data
test_df = read.csv("C:/Users/Deokinanan/Desktop/2019 -2021 CUNY SPS/FALL 2020/DATA 621/WEEK 3/moneyball-evaluation-data.csv")

@@

results
```

>>>>

---
title: "DATA 621 - HW1 - Regression Model"
author: "Samantha Deokinanan"
date: "September 3, 2020"
output:
  pdf_document: default
---

## Task

@@

## The Dataset

@@

```{r message=FALSE, warning=FALSE, include=FALSE}
# The required R packages
library(tidyverse)
library(psych)
library(olsrr)
library(rcompanion)
library(mice)
library(VIM)
library(mctest)
library(caret)

# Load training data set
theURL = "https://raw.githubusercontent.com/aadler/
DT621_Fall2020_Group2/master/HW1/data/moneyball-training-data.csv"
train_df = data.frame(read.csv(file = theURL, header = TRUE, sep = ","))
train_df = subset(train_df, select = -c(INDEX))
```

## Data Transformation, Outliers and Missing Data

@@

```{r include=FALSE}
# Quick Exploration
# sapply(train_df, describe)

@@

Next, the outlier plots revealed that there are a few extreme values that can influence the analysis. Because the objective is so create a multivariate regression model, declaring an observation as an outlier based on a just one feature could lead to unrealistic inferences. Therefore Cook's distance is use to decide if an individual entity is an extreme value or not.

Cook’s distance is a measure computed with respect to a given regression model and therefore is impacted only by the predictors included in the model. It computes the influence exerted by each data point on the response variable. Fitting the full model during this exploratory stage, the process remove the most influential points. Note from the plots that all influential observations are not necessarily outliers. Less than 3% of the data was removed.

```{r include=FALSE}
# Outlier Plots
model = lm(TARGET_WINS ~ ., data = train_df)
# ols_plot_diagnostics(model)
p1 = ols_plot_cooksd_chart(model)
p2 = ols_plot_resid_fit(model)
p3 = ols_plot_resid_hist(model)

@@

# New diagnostic plots
model_new = lm(TARGET_WINS ~ ., data = train_df)
# ols_plot_diagnostics(model_new)
p4 = ols_plot_cooksd_chart(model_new)
p5 = ols_plot_resid_fit(model_new)
p6 = ols_plot_resid_hist(model_new)
```
```{r echo=FALSE, fig.height=7}
gridExtra::grid.arrange(p1,p4,p2,p5,p3,p6, nrow=3, top = "Before and After Outliers were Removed")
```

@@

```{r echo=FALSE, fig.height=5}
aggr(train_df, col = c('darkseagreen2','darkslateblue'), numbers = TRUE, sortVars = FALSE, 
     oma = c(10,5,5,3), labels = names(train_df), cex.axis = 0.8, gap = 3, axes = TRUE, 
     Prop = TRUE, ylab = c("Proportion of Missing Data", "Combination"))

train_df.clean = complete(mice(data = train_df, method = "pmm", maxit = 5, 
                               seed = 525, print = FALSE), 3)
# sapply(train_df.clean, describe)
```

## Initial Tests

@@

```{r include=FALSE}
# Data Transformation
train_df.clean$TEAM_BATTING_1B = train_df.clean$TEAM_BATTING_H - 
  (train_df.clean$TEAM_BATTING_2B + train_df.clean$TEAM_BATTING_3B + 
     train_df.clean$TEAM_BATTING_HR)
train_df.clean = subset(train_df.clean, select = -c(TEAM_BATTING_H))

@@

```{r, fig.width=7, fig.height=7, echo=FALSE}
# Correlation
corr = round(cor(train_df.clean), 2)
corr = cbind(names = rownames(corr), data.frame(corr, row.names=NULL))
g.corr = gather(as.data.frame(corr), key, value, 2:length(corr), na.rm = TRUE)

ggplot(data = g.corr, aes(key, names, fill = value)) + geom_tile(color = "white") +
 scale_fill_gradient2(low = "darkseagreen2", high = "darkslateblue", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", name="Pearson\nCorrelation") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 12, hjust = 1)) +
  coord_fixed() + geom_text(aes(key, names, label = value), color = "black", size = 3) 
```
Lastly, because a robust regression process will be performed, higher order polynomials variables were introduced into the full model.

```{r echo=FALSE}
stepdata = cbind(train_df.clean,sapply(train_df.clean[2:length(train_df.clean)], 
                                       function(x) x^2))
names(stepdata) = make.unique(names(stepdata), sep = "_")
```

## Regression Model: Stepwise Regression with Repeated k-fold Cross-Validation

A stepwise variable selection model is conducted to determine what are the variables that can help predict the number of wins for the team. The stepwise variable selection allows variables to be added one at a time to the model, as long as the F-statistic is below the specified $\alpha$, in this case $\alpha = 0.05$. However, variables already in the model do not necessarily stay in. The steps evaluate all of the variables already included in the model and remove any variable that has an insignificant F-statistic. Only after this test ends, is the best model found, that is when none of the variables can be excluded and every variable included in the model is significant.

@@

After the steps, the final model resulted below, with $adj. R^2 = 0.43$, suggesting that this model accounts for nearly 43% of the variation in the dependent variable with the independent variables, which is acceptable as a good model. With this method of stepwise regression, AIC (Akaike Information Criteria) quantifies the amount of information loss due to simplification. That is, based on the AIC, the final model outputted is the simplest model without impacting much on the performance.

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# Set up repeated k-fold cross-validation
set.seed(525)
train.control = trainControl(method = "cv", number = 10)

@@

@@

# Summary of the model
mod_sum = summary(step.model$finalModel)

# Prediction on training data
predictions = predict(step.model, newdata = stepdata)
```

```{r echo=FALSE}
# Model accuracy
data.frame(r.squared = c(mod_sum[["r.squared"]]), adj.r.squared = c(mod_sum[["adj.r.squared"]]), 
           rsme = c(sqrt(mean((stepdata$TARGET_WINS - predictions)^2)))) %>% knitr::kable(digits = 2L)
```

## Prediction

@@

```{r echo=FALSE}
step.model$finalModel[["coefficients"]] %>% knitr::kable()
```

Using the test data and the final step model, a comparison in the prediction statistic was conducted. It is apparent that the model's prediction are not too off from the training data. However, the predictions resulted in a higher kurtosis, thus there are expectation of greater extremity of the deviations, and not centered near the mean.

```{r message=FALSE, warning=FALSE, echo=FALSE}
# load the test data
theURL = "https://raw.githubusercontent.com/aadler/
DT621_Fall2020_Group2/master/HW1/data/moneyball-evaluation-data.csv"
test_df = data.frame(read.csv(file = theURL, header = TRUE, sep = ","))

@@

results[,-c(2,8,11)] %>% knitr::kable(digits = 2L)
```
